---
title: "Introduction to Machine Learning"
author: "Ismaël Lajaaiti"
date: "06/03/2022"
output: html_document
editor_options: 
chunk_output_type: console
---

## Set up 

```{r, echo = F}
set.seed(113) # set seed for reproducibility 
```

Installing the Machine Learning framework.

```{r}
installed.pkg = installed.packages()[,1] # installed packages on your system

# If not installed, download the required packages.
required_pkg = c("torch", "torchvision", "luz") # list of required packages 
for (pkg in required_pkg) {
  if (!is.element(pkg, installed.pkg)){install.packages(pkg)}
}
```

In addition download the dataset we will be using here.

```{r}
remotes::install_github("mlverse/torchdatasets")
```

Load libraries.

```{r}
library("torch") 
library("torchvision")
library("luz")
library("torchdatasets")
```

Quick test to check if it torch was installed correctly:

```{r, collapse = T}
torch_tensor(1)
```


## Prepare the dataset

"Guess the correlation" is a dataset that asks one to guess the correlation
between two variables from their scatter plot. 

For example can you guess the correlation between the variables displayed 
in the scatter plots below?

```{r, echo = F, fig.width = 9, fig.height = 5}
par(mfrow=c(1,2)) # 1 row, 2 columns
lim = c(-0.2,1.2) # axis limits

n = 500 # number of observations

# Plot two uncorrelated variables.
X1 = runif(n)
Y1 = runif(n)
plot(X1, Y1, xlim = lim, ylim = lim, pch = 20)

# Plot two correlated variables.
X2 = runif(n)
Y2 = X2 + rnorm(n, 0, 0.1)
plot(X2, Y2, xlim = lim, ylim = lim, pch = 20)
```

And the answer was...

```{r, collapse = T}
cat("cor(X1,Y1) = ", cor(X1,Y1)) 
cat("cor(X2,Y2) = ", cor(X2,Y2)) 
```

This dataset contains contains 150,000 observations, where each observation 
contains: a scatter plot (input) and the corresponding correlation (output). 
As the dataset is large we will restrict ourselves to a subset to reduce 
the training time of the neural network.

```{r}
train_ind = 1:10000 # 10,000 observations for training 
valid_ind = 10001:15000 # 5,000 observations for validation 
test_ind  = 15001:20000 # 5,000 observations for testing 
```

Let's download the data.

```{r, collapse = T}
root      = file.path(tempdir(), "correlation")
train_set = guess_the_correlation_dataset(root = root, indexes = train_ind,
                                          download = T)
```

We will have a look at the first observation to see how it looks like.
Each observation has two parts: 
- `$x` = the input data, here the scatter plot 
- `$y` = the expected output (or the 'label'), here the correlation 

```{r}
first_element = train_set[1]
input  = first_element$x # torch tensor (150,150) = scatter plot 
output = first_element$y # torch tensor (1,) = correlation coef. 
```

The scatter plot is encoded as a tensor, but we can plot it as follow:

```{r}
input %>% as.array() %>% as.raster() %>% plot()
```

We see that the scatter plot includes the axis that are unnecessary for our 
inference task, then let's remove the axis by cropping the picture. 

```{r}
cropped_input = input[0:130, 21:150] # input without axis 
cropped_input %>% as.array() %>% as.raster() %>% plot()
```


```{r}
#b = guess_the_correlation_dataset(
# where to unpack
#root = root,
# additional preprocessing 
#transform = function(img) crop_axes(img) %>% add_channel_dim(),
# don't take all data, but just the indices we pass in
#indexes = 2:100,
#download = FALSE
#)
```


# Work with batches 

# Create the neural network 

# Train the neural network 

# Evaluate performance 



