---
title: "Introduction to Machine Learning"
author: "Ismaël Lajaaiti"
date: "06/03/2022"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
chunk_output_type: console
---

Inspired from [this tutorial](https://torch.mlverse.org/start/guess_the_correlation/).

## Set up

```{r, echo = F}
set.seed(113) # set seed for reproducibility 
```

Installing the Machine Learning framework.

```{r}
installed.pkg = installed.packages()[,1] # installed packages on your system

# If not installed, download the required packages.
required_pkg = c("torch", "torchvision") # list of required packages 
for (pkg in required_pkg) {
  if (!is.element(pkg, installed.pkg)){install.packages(pkg)}
}
```

In addition download the dataset we will be using.

```{r}
remotes::install_github("mlverse/torchdatasets")
```

Load libraries.

```{r}
library("torch") 
library("torchvision")
library("torchdatasets")
```

Quick test to check if torch is installed correctly:

```{r, collapse = T}
torch_tensor(1)
```

## Prepare the dataset

*Guess the correlation* is a dataset that asks one to guess the
correlation between two variables from their scatter plot.

For example can you guess the correlation between the variables
displayed in the scatter plots below?

```{r, echo = F, fig.width = 9, fig.height = 5}
par(mfrow=c(1,2)) # 1 row, 2 columns
lim = c(-0.2,1.2) # axis limits

n = 500 # number of observations

# Plot two uncorrelated variables.
X1 = runif(n)
Y1 = runif(n)
plot(X1, Y1, xlim = lim, ylim = lim, pch = 20)

# Plot two correlated variables.
X2 = runif(n)
Y2 = X2 + rnorm(n, 0, 0.1)
plot(X2, Y2, xlim = lim, ylim = lim, pch = 20)
```

The answer is...

```{r, collapse = T}
cat("cor(X1,Y1) = ", cor(X1,Y1)) 
cat("cor(X2,Y2) = ", cor(X2,Y2)) 
```

This dataset contains contains 150,000 observations, where each
observation contains: a scatter plot (input) and the corresponding
correlation (output). As the dataset is large, to reduce the training time of 
the neural network we will restrict ourselves to a subset.

```{r}
train_ind = 1:10000 # 10,000 observations for training 
valid_ind = 10001:15000 # 5,000 observations for validation 
test_ind  = 15001:20000 # 5,000 observations for testing 
```

Let's download the data and pick the indices for the train set.

```{r, collapse = T}
root     = file.path(tempdir(), "correlation")
train_ds = guess_the_correlation_dataset(root = root, indexes = train_ind,
                                          download = T)
```

First, we will have a look at the first observation to see how it looks like.
Each observation has two parts:

-   `$x` = the input data, here the scatter plot
-   `$y` = the expected output (or the 'label'), here the correlation

```{r}
first_element = train_ds[1]
input  = first_element$x # torch tensor (150,150) = scatter plot 
output = first_element$y # torch tensor (1,) = correlation coef. 
```

The scatter plot is encoded as a tensor, but we can plot it as follow:

```{r, fig.width = 4, fig.height = 4}
input %>% as.array() %>% as.raster() %>% plot()
```

The scatter plot includes the axis that are unnecessary to infer the
correlation, then let's remove them by cropping the picture.

```{r, fig.width = 4, fig.height = 4}
cropped_input = input[0:130, 21:150] # input without axis 
cropped_input %>% as.array() %>% as.raster() %>% plot() # plot to check
```

We have already create the training set above (`train_ds`). In addition,
we create the validation and test sets. As the data was already
download before, we do not need to download it again. We just pick
different indices.

```{r}
# Validation set.
valid_ds = guess_the_correlation_dataset(root = root, indexes = valid_ind,
    download = F)

# Test set.
test_ds = guess_the_correlation_dataset(root = root, indexes = test_ind,
    download = F)
```

Quick check of the size of our train, validation and test sets

```{r collapse=TRUE}
length(train_ds)
length(valid_ds)
length(test_ds)
```

To make the neural network training faster, we will work with *batches*.
A batch is a subset that contains a bunch of observations (here 64).
During the training, instead of going over each observation
individually, the neural network will process simultaneously the observations
belonging to the same batch. In the torch framework, this process of splitting 
each set in batches is handled by the objects called `dataloader`.

```{r}
train_dl = dataloader(train_ds, batch_size = 64, shuffle = T)
valid_dl = dataloader(valid_ds, batch_size = 64, shuffle = F)
test_dl  = dataloader(test_ds , batch_size = 64, shuffle = F)
```

The length of the dataloader is equal to its number of batches. That
latter can be recovered by dividing the number of observations in the
dataset by the batch size (here 64).

```{r collapse = T}
length(train_dl)
ceiling(length(train_ds) / 64)
```

Let's have look to the first batch.

```{r collapse = T}
first_batch <- dataloader_make_iter(train_dl) %>% dataloader_next()
dim(first_batch$x) # 64 first scatter plots (64, 150, 150)
dim(first_batch$y) # 64 first correlations (64,)
```

Our data is well-prepared and ready to be given to our neural
network. Then we can now build the neural network.

## Build the neural network

In the torch framework, a neural network is defined by two functions:

1.  `initialize` which describes the structure of the network i.e. the
    number of layers and the number of neurons per layer

2.  `forward` which tells how the data goes through the different layers
    previously defined

```{r}
torch_manual_seed(113) # set seed for reproducibility 

# Create neural network.
dnn <- nn_module(
  
  "corr-dnn", # network name 
  
  # Define layers.
  initialize = function(n_in, n_hidden, n_out) {
    self$fc1 <- nn_linear(in_features = n_in    , out_features = n_hidden)
    self$fc2 <- nn_linear(in_features = n_hidden, out_features = n_hidden)
    self$fc3 <- nn_linear(in_features = n_hidden, out_features = n_hidden)
    self$fc4 <- nn_linear(in_features = n_hidden, out_features = n_out)
  },
  
  # How input goes through the network.
  forward = function(x) {
    x %>% # input
      self$fc1() %>% # 1st layer 
      nnf_relu() %>% # activation function
 
      self$fc2() %>% # 2nd layer
      nnf_relu() %>% # activation function
 
      self$fc3() %>% # etc.
      nnf_relu() %>% # ...

      self$fc4()     # output
  }
)
```

The `initialize` function takes three arguments:

1.  `n_in`: the size of the input, here the number of pixels of the
    scatter plot, we flatten the 2d picture of size (height, width) into
    a 1d vector of size (width\*height)

2.  `n_hidden`: the number of neurons in the hidden layers

3.  `n_out`: the size of output, here 1 as the output is the correlation

Arguments can be added, deleted or replaced depending on your needs. Here is 
just a simple example that fits most situations. 

```{r}
height = dim(cropped_input)[1] # height of the picture 
width  = dim(cropped_input)[2] # width 
n_in  = height * width # total number of pixels 
n_out = 1 # expect one value in output: the correlation 
n_hidden = 100 # play with this value to optimize the network
```

We can call the function and create our neural network, and check its
structure.

```{r collapse = T}
network = dnn(n_in, n_hidden, n_out)
opt = optim_adam(params = network$parameters) # for param. opt. during training
network
# ----------------------------------------
# WEIGHTS           + BIAS    = PARAMETERS
# ----------------------------------------
# node_in*n_out     + n_out
n_in     * n_hidden + n_hidden
n_hidden * n_hidden + n_hidden
n_hidden * n_hidden + n_hidden
n_hidden * n_out    + n_out
```

The network is ready to be trained.

## Training loop

One iteration of the training loop is called an *epoch* and 
can be summarized as follow:

1.  Training over the batches of the training set
    a. Make prediction on a batch
    b. Compute the prediction error
    c. Change weight to reduce the prediction error
    d. Repeat for next batch, until the end of train set...
2.  Validation over the batches of the validation set
    a. Make prediction on a batch
    b. Compute prediction error
    d. Repeat for next batch, until the end of validation set...
3.  Early stopping check
    a. If error has decreased continue the training
    b. Else stop the training

To build that loop we specify first how the network should process 
batches during the training step...

```{r}
trainStep = function(network, batch){
  opt$zero_grad() 
  batch_size = dim(batch$x)[1]
  input = batch$x
  input = input[1:batch_size, 0:130, 21:150] # crop axis
  input = input$reshape(c(batch_size, 130*130)) # flatten input 
  output.pred = network(input)$squeeze(2) # correlation predicted by the DNN
  output.true = batch$y # expected correlation
  loss = nnf_mse_loss(output.pred, output.true) # error between pred and true
  loss$backward() # backward propagation 
  opt$step()
  loss$item() # get value 
}
```

... then during the validation step...

```{r}
validStep = function(network, batch){
  batch_size = dim(batch$x)[1]
  input = batch$x
  input = input[1:batch_size, 0:130, 21:150] # crop axis
  input = input$reshape(c(batch_size, 130*130)) # flatten input  
  output.pred = network(input)$squeeze(2) # correlation predicted by the DNN
  output.true = batch$y # expected correlation
  loss = nnf_mse_loss(output.pred, output.true) # error between pred and true
  loss$item() # get value 
}
```

... and now we can assemble the pieces.

```{r}
trainingLoop = function(network, train_dl, valid_dl,
                        epoch_max = 5, patience = 2){
  
  # Initialize variables. 
  epoch = 1
  count = 0
  loss.previous = 100
  
  # Training loop.
  while(epoch <= epoch_max & count < patience){
    
    # Train.
    network$train()
    Loss.train = c()
    coro::loop(for (batch in train_dl){
      loss.train = trainStep(network, batch) # go over the batch 
      Loss.train = c(Loss.train, loss.train) # save loss value
    })
    cat(sprintf("Epoch %0.3d/%0.3d - Train - Loss = % 3.5f \n",
                epoch, epoch_max, mean(Loss.train))) # progress
    
    # Validation.
    network$eval()
    Loss.valid = c()
    coro::loop(for (batch in valid_dl){
      loss.valid = validStep(network, batch) # go over the batch 
      Loss.valid = c(Loss.valid, loss.valid) # save loss value
    })
    cat(sprintf("Epoch %0.3d/%0.3d - Valid - Loss = % 3.5f \n",
                epoch, epoch_max, mean(Loss.valid))) # progress
    
    # Check for early stopping.
    loss.now = mean(Loss.valid) # mean loss at the current epoch
    if (loss.now > loss.previous){ # if loss has increase, increment counter 
      count = count + 1}else{ # else, reset counter and loss baseline value
      count = 0
      loss.previous = loss.now
    }
    
    epoch = epoch + 1 # next epoch 
    
  }
}
```

Training the network (it can take some times...)

```{r collapse = T}
trainingLoop(network, train_dl, valid_dl, epoch_max = 5)
```

## Evaluate the network 

Our network is now trained and ready to predict correlations from scatter plots 
it has *never* seen. That what the test set is for. 
Let's specify how the network is tested on one batch...

```{r}
testStep = function(network, batch){
  batch_size = dim(batch$x)[1]
  input = batch$x
  input = input[1:batch_size, 0:130, 21:150] # crop axis
  input = input$reshape(c(batch_size, 130*130)) # flatten input  
  output.pred = network(input)$squeeze(2) # correlation predicted by the DNN
  output.true = batch$y # expected correlation
  output = list(pred = output.pred, true = output.true)
  return(output)
}
```

... and then loop over all the batches of the test set.

```{r}
testLoop = function(network, test_dl){
  Pred = c() # store predicted correlations
  True = c() # store true correlations
  coro::loop(for (batch in test_dl){
      output = testStep(network, batch) # go over the batch 
      Pred = c(Pred, output$pred %>% as.array()) # save pred. values
      True = c(True, output$true %>% as.array()) # save true values 
    })
  out = list(true = True, pred = Pred)
  return(out)
}
```

Testing the network.

```{r}
out = testLoop(network, test_dl)
```

The most direct way to have a quick idea of how well our network performs is to 
plot the predicted correlations against the true correlations.

```{r, fig.width = 7, fig.height = 7}
# Plot Pred vs. True. 
plot(out$true, out$pred, xlab = 'True correlation',
     ylab = 'Predicted correlation', cex = 0.5, col = rgb(.1,.1,.1, 0.6))
# Plot 1:1 line.
abline(0, 1, lw = 4)
# Add legend. 
legend("topleft", legend = c("Data", "1:1"), lwd = c(NA, 4), pch = c(1, NA))
```

The performances of the network can be quantified more accurately
(e.g. by decomposing the total error in variance and bias),
but that's beyond the scope of this introduction.
