---
title: "Introduction to Machine Learning"
author: "Ismaël Lajaaiti"
date: "06/03/2022"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
chunk_output_type: console
---

## Set up

```{r, echo = F}
set.seed(113) # set seed for reproducibility 
```

Installing the Machine Learning framework.

```{r}
installed.pkg = installed.packages()[,1] # installed packages on your system

# If not installed, download the required packages.
required_pkg = c("torch", "torchvision", "luz") # list of required packages 
for (pkg in required_pkg) {
  if (!is.element(pkg, installed.pkg)){install.packages(pkg)}
}
```

In addition download the dataset we will be using here.

```{r}
remotes::install_github("mlverse/torchdatasets")
```

Load libraries.

```{r}
library("torch") 
library("torchvision")
library("luz")
library("torchdatasets")
```

Quick test to check if it torch was installed correctly:

```{r, collapse = T}
torch_tensor(1)
```

## Prepare the dataset

"Guess the correlation" is a dataset that asks one to guess the
correlation between two variables from their scatter plot.

For example can you guess the correlation between the variables
displayed in the scatter plots below?

```{r, echo = F, fig.width = 9, fig.height = 5}
par(mfrow=c(1,2)) # 1 row, 2 columns
lim = c(-0.2,1.2) # axis limits

n = 500 # number of observations

# Plot two uncorrelated variables.
X1 = runif(n)
Y1 = runif(n)
plot(X1, Y1, xlim = lim, ylim = lim, pch = 20)

# Plot two correlated variables.
X2 = runif(n)
Y2 = X2 + rnorm(n, 0, 0.1)
plot(X2, Y2, xlim = lim, ylim = lim, pch = 20)
```

And the answer was...

```{r, collapse = T}
cat("cor(X1,Y1) = ", cor(X1,Y1)) 
cat("cor(X2,Y2) = ", cor(X2,Y2)) 
```

This dataset contains contains 150,000 observations, where each
observation contains: a scatter plot (input) and the corresponding
correlation (output). As the dataset is large we will restrict ourselves
to a subset to reduce the training time of the neural network.

```{r}
train_ind = 1:10000 # 10,000 observations for training 
valid_ind = 10001:15000 # 5,000 observations for validation 
test_ind  = 15001:20000 # 5,000 observations for testing 
```

Let's download the data.

```{r, collapse = T}
root     = file.path(tempdir(), "correlation")
train_ds = guess_the_correlation_dataset(root = root, indexes = train_ind,
                                          download = T)
```

We will have a look at the first observation to see how it looks like.
Each observation has two parts:

-   `$x` = the input data, here the scatter plot
-   `$y` = the expected output (or the 'label'), here the correlation

```{r}
first_element = train_ds[1]
input  = first_element$x # torch tensor (150,150) = scatter plot 
output = first_element$y # torch tensor (1,) = correlation coef. 
```

The scatter plot is encoded as a tensor, but we can plot it as follow:

```{r, fig.width = 4, fig.height = 4}
input %>% as.array() %>% as.raster() %>% plot()
```

We see that the scatter plot includes the axis that are unnecessary for
our inference task, then let's remove the axis by cropping the picture.

```{r, fig.width = 4, fig.height = 4}
cropped_input = input[0:130, 21:150] # input without axis 
cropped_input %>% as.array() %>% as.raster() %>% plot()
```

We have already built the training set above. In addition, we build the 
validation and test dataset. As the data was already download before, we 
do not need to download it again. We just pick different indices. 

```{r}
valid_ds = guess_the_correlation_dataset(root = root, indexes = valid_ind,
    download = F)

test_ds = guess_the_correlation_dataset(root = root, indexes = test_ind,
    download = F)
```

Quick check of the size of our train, validation and test sets

```{r collapse=TRUE}
length(train_ds)
length(valid_ds)
length(test_ds)
```


```{r}
train_dl = dataloader(train_ds, batch_size = 64, shuffle = T)
valid_dl = dataloader(valid_ds, batch_size = 64, shuffle = F)
test_dl  = dataloader(test_ds , batch_size = 64, shuffle = F)
```


```{r}
length(train_dl)
ceiling(length(train_ds) / 64)
```


```{r}
first_batch <- dataloader_make_iter(train_dl) %>% dataloader_next()
dim(first_batch$x) # 64 first scatter plots (64, 150, 150)
dim(first_batch$y) # 64 first correlations (64,)
```

  
```{r}
torch_manual_seed(113) # set seed for reproducibility 

# Create neural network.
dnn <- nn_module(
  
  "corr-dnn", # network name 
  
  # Define hidden layers.
  initialize = function(n_in, n_hidden, n_out) {
    self$fc1 <- nn_linear(in_features = n_in    , out_features = n_hidden)
    self$fc2 <- nn_linear(in_features = n_hidden, out_features = n_hidden)
    self$fc3 <- nn_linear(in_features = n_hidden, out_features = n_hidden)
    self$fc4 <- nn_linear(in_features = n_hidden, out_features = n_out)
  },
  
  # How input goes through the network.
  forward = function(x) {
    x %>% # input
      self$fc1() %>% # 1st layer 
      nnf_relu() %>% # activation function
 
      self$fc2() %>% # 2nd layer
      nnf_relu() %>% # activation function
 
      self$fc3() %>% # etc.
      nnf_relu() %>% # ...

      self$fc4()     # output
  }
)
```


```{r}
height = dim(cropped_input)[1] # height of the picture 
width  = dim(cropped_input)[2] # width 
n_in  = height * width # total number of pixels 
n_out = 1 # expect one value in output: the correlation 
n_hidden = 100 # play with this value to optimize the network
```


```{r}
network = dnn(n_in, n_hidden, n_out)
opt = optim_adam(params = network$parameters) # for param. opt. during training
network
# WEIGHTS           + BIAS    = PARAMETERS
# node_in*n_out     + n_out
n_in     * n_hidden + n_hidden
n_hidden * n_hidden + n_hidden
n_hidden * n_hidden + n_hidden
n_hidden * n_out    + n_out
```


```{r}
trainStep = function(network, batch){
  opt$zero_grad() 
  batch_size = dim(batch$x)[1]
  input = batch$x
  input = input[1:batch_size, 0:130, 21:150] # crop axis
  input = input$reshape(c(batch_size, 130*130)) # flatten input 
  output.pred = network(input)$squeeze(2) # correlation predicted by the DNN
  output.true = batch$y # expected correlation
  loss = nnf_mse_loss(output.pred, output.true) # error between pred and true
  loss$backward() # backward propagation 
  opt$step()
  loss$item() # get value 
}
```


```{r}
validStep(network, first_batch)
```


```{r}
validStep = function(network, batch){
  batch_size = dim(batch$x)[1]
  input = batch$x
  input = input[1:batch_size, 0:130, 21:150] # crop axis
  input = input$reshape(c(batch_size, 130*130)) # flatten input  
  output.pred = network(input)$squeeze(2) # correlation predicted by the DNN
  output.true = batch$y # expected correlation
  loss = nnf_mse_loss(output.pred, output.true) # error between pred and true
  loss$item() # get value 
}
```


```{r}
trainingLoop = function(network, train_dl, valid_dl,
                        epoch_max = 5, patience = 2){
  
  # Initialize variables. 
  epoch = 1
  count = 0
  loss.previous = 100
  
  # Training loop.
  while(epoch < epoch_max & count < patience){
    
    # Train.
    network$train()
    Loss.train = c()
    coro::loop(for (batch in train_dl){
      loss.train = trainStep(network, batch) # go over the batch 
      Loss.train = c(Loss.train, loss.train) # save loss value
    })
    cat(sprintf("Epoch %0.3d/%0.3d - Train - Loss = % 3.5f \n",
                epoch, epoch_max, mean(Loss.train))) # progress
    
    # Validation.
    network$eval()
    Loss.valid = c()
    coro::loop(for (batch in valid_dl){
      loss.valid = validStep(network, batch) # go over the batch 
      Loss.valid = c(Loss.valid, loss.valid) # save loss value
    })
    cat(sprintf("Epoch %0.3d/%0.3d - Valid - Loss = % 3.5f \n",
                epoch, epoch_max, mean(Loss.valid))) # progress
    
    # Check for early stopping.
    loss.now = mean(Loss.valid) # mean loss at the current epoch
    if (loss.now > loss.previous){ # if loss has increase, increment counter 
      count = count + 1}else{ # else, reset counter and loss baseline value
      count = 0
      loss.previous = loss.now
    }
    
    epoch = epoch + 1 # next epoch 
    
  }
}
```


```{r}
trainingLoop(network, test_dl, valid_dl)
```

```{r}
testStep = function(network, batch){
  batch_size = dim(batch$x)[1]
  input = batch$x
  input = input[1:batch_size, 0:130, 21:150] # crop axis
  input = input$reshape(c(batch_size, 130*130)) # flatten input  
  output.pred = network(input)$squeeze(2) # correlation predicted by the DNN
  output.true = batch$y # expected correlation
  output = list(pred = output.pred, true = output.true)
  return(output)
}
```


```{r}
testLoop = function(network, test_dl){
  Pred = c() # store predicted correlations
  True = c() # store true correlations
  coro::loop(for (batch in test_dl){
      output = testStep(network, batch) # go over the batch 
      Pred = c(Pred, output$pred %>% as.array()) # save pred. values
      True = c(True, output$true %>% as.array()) # save true values 
    })
  out = list(true = True, pred = Pred)
  return(out)
}
```

```{r}
out = testLoop(network, valid_dl)
```

```{r}
plot(out$true, out$pred)
abline(0,1)
```



# Work with batches

# Create the neural network

# Train the neural network

# Evaluate performance
